---
title: "OA and GSEA"
author: "Rebekah Sheih"
date: "2026-04-29"
output:
  html_document:
    fig_width: 8
    fig_height: 5
    dpi: 200
---

# Enrichment analysis of the full SMARCE1 dataset

Up until now, we have been working with the SMARCE1 dataset using reads that only align to chromosome 17. This was done to speed things up.

Now, we would like to run DeSeq2 and perform Gene Ontology on the full dataset. 

I ran Salmon on the full dataset fastq files for you- so you can start with the salmon quant.sf files.

Below is a pipeline for analyzing differential expression and performing Gene Ontology (GO) enrichment analysis with DESeq2, tximport, and clusterProfiler.

You will need to run on your laptop in R studio (too big for cloud).

Running this yourself, you might encounter some real world difficulties in installing packages. That's real life, and you'll likely encounter similar challenges for your independent project.


```{r setup4, quiet=TRUE, echo=FALSE, message=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org"))

if (!require("tximport")) {
  BiocManager::install("tximport")}

if (!require("readr")) {
  install.packages("readr")}

if (!require("ggplot2", quietly = TRUE))
  install.packages("ggplot2")

if (!require("DESeq2", quietly = TRUE))
  BiocManager::install("DESeq2")

if (!require("ggrepel", quietly = TRUE))
  BiocManager::install("ggrepel")

if (!require("tximport", quietly = TRUE))
  BiocManager::install("tximport")

if (!require("clusterProfiler", quietly = TRUE))
  BiocManager::install("clusterProfiler")

if (!require("apeglm", quietly = TRUE))
  BiocManager::install("apeglm")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")


```

```{r}
if (!require("vsn")) {
  BiocManager::install("vsn")}
```


```{r lib, quiet=TRUE, echo=FALSE, message=FALSE}
library(tximport)
library(readr)
library(here)
library(DESeq2)
library(ggrepel)
library(tidyverse)
library(clusterProfiler)
library(apeglm)
library(dplyr)
library(vsn)
```

Here we read in the metadata for the full dataset
```{r}
# samples <- read.csv("meta_smarce1.csv", header = TRUE)
# samples$Run

samples <- data.frame(
  Run = c(
    "NZXTV9_1_U2OS_rep_1_trim.fastq.gz",
    "NZXTV9_2_U2OS_rep_2_trim.fastq.gz",
    "NZXTV9_3_U2OS_rep_3_trim.fastq.gz",
    "NZXTV9_4_PMLKO_rep_1_trim.fastq.gz",
    "NZXTV9_5_PMLKO_rep_2_trim.fastq.gz",
    "NZXTV9_6_PMLKO_rep_3_trim.fastq.gz"
  ),
  condition = c("WT", "WT", "WT", "PMLKO", "PMLKO", "PMLKO")
)
```

# import salmon count files.
```{r}

# we import the salmon count files in the directory "salmon_quant"

salmon_dir <- here("salmon_quant")

```
We use file.path to build a file path for each of the samples "quant.sf" data.
```{r, quiet=TRUE, echo=FALSE, message=FALSE}
# files <- file.path(here("salmon"), samples$Run, "quant.sf")
files <- list.files(salmon_dir, pattern = "quant.sf", recursive = TRUE, full.names = TRUE)
print(files)
file.exists(files) # check if file exists
```

Now, we add names to the vector of file paths.
```{r}
names(files) <- samples$Run
files

#this is important b/c now we can refer to the files by name instead of just position in the list.

files["NZXTV9_2"]
```
Build tx2gene.tsv

```{r}
if (!require("GenomicFeatures")) {
  BiocManager::install("GenomicFeatures")}

if (!require("biomaRt")) {
  BiocManager::install("biomaRt")}
```

```{r}
library(GenomicFeatures)

txdb <- makeTxDbFromGFF("/Users/rebekahsheih/Desktop/pxl_rnaseq/gencode.v49.annotation.gtf") 

k <- keys(txdb, keytype = "TXNAME")

tx2gene <- select(txdb,
                  keys = k,
                  columns = c("TXNAME", "GENEID"),
                  keytype = "TXNAME")

library(biomaRt)

ensembl <- useEnsembl("genes", dataset = "hsapiens_gene_ensembl")

gene_names <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name"),
  mart = ensembl
)

tx2gene <- merge(tx2gene, gene_names,
                 by.x = "GENEID",
                 by.y = "ensembl_gene_id",
                 all.x = TRUE)

```

Next we import a table that contains the transcript to gene name mapping (like before).
```{r}
# tx2gene <- read.delim("tx2gene.tsv", header = FALSE, stringsAsFactors = FALSE) # ensmbl gene names
tx2gene <- read.delim("tx2gene.csv", header = FALSE, stringsAsFactors = FALSE) # ensmbl gene names

tx2gene <- tidyr::separate(tx2gene,
                            col = V1,
                            into = c("TXNAME", "GENEID"),
                            sep = ",")
tx2gene <- tx2gene[-1, ]

head(tx2gene)

```


Now we use tximport to load in the Salmon results and covert transcript-level estimates to gene-level expression.

txi will be a variable that is a list. It is like having a bunch of spreadsheets (one per sample) loaded into a master table. 

```{r}
txi <- tximport(files, type="salmon", tx2gene=tx2gene, ignoreTxVersion = TRUE)

head(txi$counts)

```

Now we create the deseq2 object
```{r}
ddsTxi <- DESeqDataSetFromTximport(txi,
                                   colData = samples,
                                   design = ~ condition)
```

<!-- Warning in DESeqDataSet(se, design = design, ignoreRank) : -->
<!--   some variables in design formula are characters, converting to factors -->
<!-- using counts and average transcript lengths from tximport -->


We will prefilter out low-count genes (remove the noise)
```{r prefilter}


smallestGroupSize <- 3
keep <- rowSums(counts(ddsTxi) >= 50) >= smallestGroupSize # how many have more than 50 reads across whole thing


# keep is a logical vector (TRUE/FALSE)

sum(keep)
sum(!keep)
table(keep)

dds <- ddsTxi[keep,]

rm(ddsTxi)

```

Running Deseq2
```{r}
# factor levels

dds$condition <- relevel(dds$condition, ref = "WT")

# deseq2 reports other condition vs reference condition. 
# so log2 fold change >0 means higher in mutant vs WT.

dds <- DESeq(dds)     # this is where Deseq2 actually runs the statistics
res <- results(dds)   # extracts results into a table.


```

## Log fold change shrinkage for visualization and ranking

This section focuses on refining differential expression results. lfcShrink() applies log fold change (LFC) shrinkage to reduce noise in gene expression estimates, using the apeglm method for improved accuracy. The resultsNames() call shows available comparison coefficients. 

```{r}

resultsNames(dds)

resLFC <- lfcShrink(dds, coef="condition_PMLKO_vs_WT", type="apeglm")
resLFC


# MA plot of the data

plotMA(res, ylim=c(-2,2))

# Shrunk log2 fold changes MA plot

plotMA(resLFC, ylim=c(-2,2))

pdf("./out/plots/comparison_coefficients.pdf")

```


We count significant differentially expressed genes at different thresholds (padj < 0.1, < 0.05). Finally, we recompute results at a stricter alpha (0.05) to identify high-confidence DEGs.

```{r}

#how many adj pvalues less than 0.1?
sum(res$padj < 0.1, na.rm=TRUE)

sum(res$padj < 0.05, na.rm=TRUE)

# you can change the p value cutoff using alpha

res05 <- results(dds, alpha=0.05)
summary(res05)

sum(res05$padj < 0.05, na.rm = TRUE)


```


Let's remind ourselves what is in the res dataframe. 

```{r}

dim(res)
colnames(res)
nrow(res)

res[1,]
res[12900:12906,]
```

Now let's plot the smallest P value DEG.

```{r Counts_example, echo=FALSE}

gene_min <- which.min(res$padj)
gene_min

plotCounts(dds, gene=gene_min, intgroup="condition")
```

```

But we like common gene names in our title!

```{r}
# add gene names to the Deseq2 object.
rowData(dds)$gene_name <- tx2gene$V3[match(rownames(dds), tx2gene$V2)]


plotCounts(dds,
           gene = gene_min,
           intgroup = "condition",
           main = rowData(dds)$gene_name[gene_min])

```
```{r}
# merge transcript count to gene ID using tximport
```


Export a padj ordered list.

```{r other}
mcols(res)$description

# adj p values 

resOrdered <- res[order(res$padj),] 

# add gene name as second column
resOrdered$gene_name <- tx2gene$V3[
  match(rownames(resOrdered), tx2gene$V2)
]

#resOrdered <- resOrdered[, c("gene_name", setdiff(names(resOrdered), "gene_name"))]

head(resOrdered)

# Exporting results to CSV files

dir.create(here("out"), showWarnings = F)

write.csv(as.data.frame(resOrdered), 
          file=here("out", "condition_treated_results.csv"))

```

Extracting Transformed Values for Visualization

## Normalized, variance-stabilized transformed counts for visualization

Raw RNA-seq counts:
-have huge differences in scale
-have variance that increases with mean (heteroskedasticity)

Example:
-low-count genes → noisy
-high-count genes → very stable

This can make plots like PCA, clustering, or heatmaps misleading.


```{r normalization}
# these are for visualization and exploratory analysis. not for DE testing.

# Variance Stabilizing Transformation
vsd <- vst(dds, blind=FALSE)

head(assay(vsd), 3)
```


PCA Plot

```{r PCA, echo=FALSE}
plotPCA(vsd, intgroup="condition")
pdf("./out/plots/pca_var.pdf")
```

PCA Plot using ggplot

```{r PCA_ggplot, echo=FALSE}
dat <- plotPCA(vsd,returnData=TRUE,intgroup=c("condition"))

p <- ggplot(dat,aes(x=PC1,y=PC2,col=paste(condition)))
p <- p + geom_point() + 
  xlab(paste("PC1: ", round(attr(dat,"percentVar")[1],2)*100, "% variation explained", sep="")) + 
  ylab(paste("PC2: ", round(attr(dat,"percentVar")[2],2)*100, "% variation explained", sep="")) +
  geom_label_repel(aes(label=name))
p
```

Note that H9-5 and H9-6 might be outliers.

## Quality Control Visualization and Heatmaps

This section performs quality control visualization. 

- meanSdPlot() checks variance stability across expression levels

- the heatmap displays the top 20 expressed genes (using pheatmap). The heatmap uses normalized counts (ntd) and groups samples by condition. 


```{r moreplots, echo=F}

library("vsn")

# SD plot of the non-vsd normalized data
meanSdPlot(counts(dds, normalized=TRUE))

# SD plot of vsd normalized data
meanSdPlot(assay(vsd))

pdf("./out/plots/mean_sd_plot.pdf")

# takeaway: Before transformation, variance explodes with expression level—after VST, it’s stabilized.
```

## Heatmap of the count matrix

```{r}
library("pheatmap")

# here we will plot the 20 highest expressed genes

# Check and ensure select is valid
select <- order(rowMeans(counts(dds, normalized=TRUE)), decreasing=TRUE)[1:20]
select <- select[!is.na(select)]  # Remove NA values if present

# Ensure annotation dataframe is properly formatted
df <- as.data.frame(colData(dds)[, "condition", drop = FALSE])
colnames(df) <- "Condition"  # Ensure column name is valid

pdf("./out/plots/heatmap2.pdf")

# Generate heatmap
pheatmap(assay(vsd)[select, , drop = FALSE],  # Ensure non-empty matrix
         cluster_rows = FALSE, 
         show_rownames = TRUE,
         cluster_cols = FALSE, 
         annotation_col = df)



```

Now for most significant DEGs, plus show gene names

```{r}
select2 <- head(order(res$padj), 20)

mat <- assay(vsd)[select2, ]
# take a look at mat with View

# map ENSG → gene names
gene_names <- tx2gene$V3[match(rownames(mat), tx2gene$V2)]

# replace rownames
rownames(mat) <- gene_names

pdf("./out/plots/heatmap.pdf")

pheatmap(mat,
         cluster_rows = TRUE, 
         show_rownames = TRUE,
         cluster_cols = TRUE, 
         annotation_col = df)

```


## Volcano plot

A basic Volcano plot

```{r volcano, echo=FALSE}
# negative log-scaled adjusted p-values

log_padj <- -log10(resLFC$padj)
log_padj[log_padj > 100] <- 100

pdf("./out/plots/volcano2.pdf")

plot(resLFC$log2FoldChange,
     log_padj,
     pch = 20,
     cex = 0.5,
     col = ifelse(resLFC$padj < 0.05 & abs(resLFC$log2FoldChange) > 1,
             "red", "grey"),
     ylab = "-log10(adjusted p-value)",
     xlab = "shrunken log2 fold changes")
abline(h = -log10(0.1), col = "blue", lty = 2)
abline(v = c(-1, 1), col = "blue", lty = 2)
```

```{r}
log_padj <- -log10(resLFC$padj)
log_padj[log_padj > 100] <- 100

# define colors
cols <- ifelse(resLFC$padj < 0.1 & resLFC$log2FoldChange > 1, "red",
        ifelse(resLFC$padj < 0.1 & resLFC$log2FoldChange < -1, "blue",
               "black"))

plot(resLFC$log2FoldChange,
     log_padj,
     pch = 20,
     cex = 0.5,
     col = cols,
     ylab = "-log10(adjusted p-value)",
     xlab = "shrunken log2 fold changes")

# threshold lines
abline(h = -log10(0.1), col = "darkgrey", lty = 2)
abline(v = c(-1, 1), col = "darkgrey", lty = 2)
```



# GO analysis

Here we are using some analysis based on the CBC tutorial on GO analysis

https://github.com/CBC-UCONN/RNA-seq-with-reference-genome-and-annotation/tree/master/r_analysis

```{r GO, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


# create an object for the Ensembl Genes v100 mart

if (!requireNamespace("biomaRt", quietly = TRUE)) {
  install.packages("BiocManager")
  BiocManager::install("biomaRt")
}
library(biomaRt)

ensembl <- useEnsembl(
  biomart = "genes",
  dataset = "hsapiens_gene_ensembl",
  mirror = "useast"
)


# get gene names

ann <- getBM(
  filters = "ensembl_gene_id",
  values = rownames(res),
  attributes = c(
    "ensembl_gene_id",
    "description",
    "external_gene_name"
  ),
  mart = ensembl
)

```

```{r}
# Try if ensembl not working  
ann <- readRDS("/Users/rebekahsheih/Desktop/pxl_rnaseq/pxl_rnaseq/ann_human_ensembl_gene_ids.rds")

```

GO analysis terms


```{r}
# get GO term info
# each row is a single gene ID to GO ID mapping, so the table has many more rows than genes in the analysis

# go_ann <- getBM(filter="ensembl_gene_id",value=rownames(res),attributes=c("ensembl_gene_id","description","go_id","name_1006","definition_1006","namespace_1003"),mart=ensembl)
# this takes a while.


go_ann <- readRDS("/Users/rebekahsheih/Desktop/pxl_rnaseq/pxl_rnaseq/go_ann_human_ensembl_gene_ids.rds")


#let's look at the annotation:
filter(ann, description!="") %>% head()

head(ann)
# and Go term:

head(go_ann)

# save go_genes object.
save(go_ann, file = here::here("out", "go_ann.RData"))


# you can later load this go_annotation object for future analyses
# go_ann <- load ("./out/go_ann.RData")

###
```

```{r}
# put results and gene info in the same table
res_data <- res %>% 
  as.data.frame() %>%
  rownames_to_column(var = "ensembl_gene_id") 

head(res_data)

res_ann <- merge(x=res_data,y=ann, by.x="ensembl_gene_id", by.y="ensembl_gene_id", all.x=TRUE)

head(res_ann)

# rearrange columns
res_ann <- dplyr::select(res_ann, external_gene_name, everything())

# write to .csv
write.csv(as.data.frame(res_ann), 
          file=here::here("out","res_ann.csv"), row.names=F)

```

## Over-representation Analysis (OR) 

ORA (over-representation analysis) tests whether a GO category contains more DE genes than expected by chance, using a predefined DE gene list


```{r}


# get ENSEMBL gene IDs for genes with padj < 0.1
genes <- rownames(res[which(res$padj < 0.1),])

# get ENSEMBL gene IDs for universe (all genes with non-NA padj passed independent filtering)

univ <- rownames(res[!is.na(res$padj),])
# this will get us only genes that have adjusted p values. it is for a BACKGROUND SET OF GENES TO COMPARE TO instead of using all genes in the genome.


# pull out the columns of go_ann containing GO IDs and descriptions, keep only unique entries. 

# This checks whether your gene set (genes) is statistically enriched for any GO terms compared to the universe, based on your custom mapping in TERM2GENE.


go_ann[,c(3,4)] %>% head()
gonames <- unique(go_ann[,c(3,4)]) %>% unique()
nrow(gonames)

res_enrich <- enricher(
  gene=genes,
  universe=univ,
  TERM2GENE=go_ann[,c(3,1)],
  TERM2NAME=gonames
)


# as.data.frame(res_enrich) %>% head()

# GeneRatio: number of DE genes in category / total DE genes
# BgRation: total genes in category / total genes considered
# pvalue: raw p-value
# p.adjust fdr adjusted p-value (BH method)
# qvalue: alternative fdr adjusted p-value
```

### Enrichment Analysis Plot

```{r enrichR, echo=F}
#visuals

pdf("./out/plots/dotplot.pdf")

dotplot(res_enrich)

```

## Dotplot by GO Category

There are three GO Categories:
- Molecular Function (MF)
- Biological Process (BP)
- Cellular Component (CC)

We now add category info to your results.

```{r}
# Get unique GO ID to namespace mapping
gonamespace <- unique(go_ann[, c("go_id", "namespace_1003")])
colnames(gonamespace) <- c("ID", "Ontology")  # Rename to match enrichResult@result column

# Add namespace info to the enrichment result
res_enrich@result <- merge(res_enrich@result, gonamespace, by = "ID", all.x = TRUE)
```

Now print to PDF. 

```{r}
library(ggplot2)


pdf(here::here("out", "Enrich_Category.pdf"), width = 10, height = 10)

dotplot(res_enrich, showCategory=30) + 
  facet_wrap(~Ontology, scales = "free_y")

dev.off()

```



## Gene-set enrichment analysis (GSEA)

For explanation of overrepresentation analysis and gene set enrichment analysis see this nice video! 

[GSEA](https://www.youtube.com/watch?v=egO7Lt92gDY)

Pathway enrichment analysis (overrepresentation analysis) really depends a lot on p value cutoffs.

GSEA is different. The input is not a list of genes that are DE. It is a rank of genes. 

Ranking = SIGN(FC)* - log10(p-value)

We will determine if functional categories tend to be enriched for higher or lower expression levels

Shrunken or raw log2fc depends on preference for smoothing noise but reducing signal

Extract log2 fold changes, only for genes that passed independent filtering.


```{r gsea, echo=FALSE}

l2fcs <- as.data.frame(res) %>% 
  dplyr::filter(!is.na(padj)) %>%
  dplyr::select(log2FoldChange)

# View(l2fcs)

# put log2FCs in a vector, add gene IDs as names, sort 

l2fcvec <- l2fcs[,1]
names(l2fcvec) <- rownames(l2fcs)
# View(l2fcvec)

l2fcvec <- sort(l2fcvec, decreasing=TRUE)
# View(l2fcvec)

res_gsea <- GSEA(
  geneList=l2fcvec,
  TERM2GENE=go_ann[,c(3,1)],
  TERM2NAME=gonames
) 

```

The results table (res_gsea@result) is sorted by normalized enrichment score (NES), so:

- The first row (e.g., geneSetID = 1) is the most enriched term
    
Enrichment plot for top enriched pathway term.

```{r}
gsea_df <- as.data.frame(res_gsea@result)

gsea_sorted <- gsea_df %>%
  arrange(desc(NES))

head(gsea_sorted)

```

```{r gseaplot_top, echo=FALSE}

# pick top upregulated category
top_up <- gsea_sorted %>%
  filter(NES > 0, p.adjust < 0.05) %>%
  slice(1)

# pick top downregulated category
top_down <- gsea_sorted %>%
  arrange(NES) %>%
  filter(NES < 0, p.adjust < 0.05) %>%
  slice(1)

# plot top upregulated category
gseaplot(res_gsea,
         geneSetID = top_up$ID,
         title = top_up$Description)

# plot top downregulated category
gseaplot(res_gsea,
         geneSetID = top_down$ID,
         title = top_down$Description)

```





DotPlot of GSEA

```{r plotGSEA, echo=F}

dotplot(res_gsea) 
dotplot(res_gsea, split = ".sign") + facet_grid(. ~ .sign)


```


```{r plotGSEA_plot, echo=F}

dotplot(res_gsea) 

dotplot(res_gsea, split = ".sign") + facet_grid(. ~ .sign)

p<-dotplot(res_gsea, split = ".sign") + facet_grid(. ~ .sign)

ggplot2::ggsave("./out/gsea_dotplot.pdf",
                plot = p,
                width = 9,
                height = 10)


```


```{r}

# plot a category you are interested in.
# GO:0030182
# neuron differentiation

gseaplot(res_gsea,
         geneSetID = "GO:0030182",
         title = res_gsea@result$Description[res_gsea@result$ID == "GO:0030182"]
         )

```


## Summary

What's the difference between ORA and GSEA?

- ORA (over-representation analysis) tests whether a GO category contains more DE genes than expected by chance, using a predefined DE gene list (e.g., padj < 0.05).

- GSEA considers all genes ranked by their association with the phenotype (e.g., by log2FC), testing whether genes in a GO category are non-randomly distributed in this ranked list.
